$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-03-26 write_up/ms.tex
finished draft of Cauchy section

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement. MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care. A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior. Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling. Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use. We repeat here a quote from Chris Holmes to this effect \cite{holmes},

@@

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \cite{lambert2018Student}. MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions). In the face of only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often practically converge to the posterior in only a (relatively small) finite number of samples.

The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler. Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred. Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged. Like furious bees trapped in a room of a house \cite{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas. With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.

To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \cite{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \cite{gelman2013bayesian}. If these variances are similar, $\hat{R}\approx 1$, and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \cite{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \cite{vehtari2019rank} and adopted \cite{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that generated each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \cite{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \cite{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test). It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank}. Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.

%$R^*$ calculation is enabled in Pints inference software \cite{Clerx2019Pints}.

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.

@@

The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for models with 1000 samples), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

@@

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} and two additional examples based on a multivariate normal model.

@@

First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \cite{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \cite{vehtari2019rank}.

@@

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that the variable importance measure provides information complementary to $\hat{R}$ and ESS.

@@

\subsection{Infinite variance: Cauchy example}
We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance. Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
%
\begin{equation}
x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50,
\end{equation}
%
We call this parameterisation the ``nominal'' version of this model.

In addition, we also use Stan to sample from an ``alternative'' parameterisation of the Cauchy, based on a scale mixture of Gaussians \cite{vehtari2019rank},
%
\begin{align}
a_j \sim  \N(0,1), \qquad
b_j \sim  \text{Gamma}(0.5, 0.5), \qquad
x_j =  a_j/\sqrt{b_j}.
\end{align}
%
The distribution of the $x$ vector is the same under both parameterisations, although the thin-tailed $(a,b)$ vectors define a higher dimensional posterior that improves sampling efficiency.

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

@@

To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how close the sampling distribution is to the target. One measure of distributional "closeness" is the KL-divergence, that in this case could measure the divergence from the target to the sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated. The trouble is, for distributions like the Cauchy that have fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails. This means that approximate KL-divergence is unreliable for these types of model. Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-derived quantities. Specifically, we calculate the $R^2$ for the regression of the actual quantile values on the sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities. In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 $x_j$.

In the left-hand panel of Fig. \ref{fig:cauchy_convergence}, we plot this ``quantile $R^2$'' as a function of MCMC sample size for both the nominal and alternative parameterisations of the Cauchy model. This shows that after only c.10,000 iterations the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles. In the middle and right panels of Fig. \ref{fig:cauchy_convergence}, we show how the maximum value $\hat{R}$ across all 50 dimensions of the target and $R^*$ as calculated by Algorithm \ref{alg:R_star} (both measures are calculated on chains split into equal halves) perform as the number of MCMC samples increases. After c.550 iterations, the alternative parameterisation achieves split-$\hat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum split-$\hat{R}$ value: in both cases, these suggest convergence. By contrast, after 25,000 iterations, $R^*\approx 1.05$ for the nominal model, and $R^*>2$ for the nominal parameterisation: in both cases, these $R^*$ values suggest lack of convergence.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
	\caption{\textbf{Measuring convergence with MCMC sample size for Cauchy model.} The left-hand panel shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; the middle panel, shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; the right shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:cauchy_convergence}
\end{figure}

\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.

The model can be parameterised two ways, as described in \cite{vehtari2019rank}. The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
%
\begin{align*}
\theta_j &\sim \N(\mu, \tau), \\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
The ``non-centered'' parameterisation recodes this model in a way which does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$. This can be written as,
%
\begin{align*}
\tilde{\theta}_j &\sim \N(0, 1), \\
\theta_j &= \mu + \tau \tilde{\theta}_j,\\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
of the distribution of these effects. In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.

We first used Stan \cite{carpenter2017stan} to sample from the centered model using 4 chains. Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \cite{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased. We also used Stan with the same algorithm settings to sample from the non-centered model.

To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \cite{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains. Next, we trained the GBM on a training set comprising 2800 samples, and used it to classify the causative chain for 1200 independent test samples. Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models. The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A. In this plot, it is clear that whereas the centered model shows signs of convergence, the non-centered model does not: intuitively, the chains have not sufficiently mixed with themselves nor the others meaning that the causative chain can be predicted with reasonable accuracy from the sample location.

>>>>

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} and the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement. MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care. A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior. Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling. Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use. We repeat here a quote from Chris Holmes to this effect \citep{holmes},

@@

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}. MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions). In the face of only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often practically converge to the posterior in only a (relatively small) finite number of samples.

The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler. Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred. Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged. Like furious bees trapped in a room of a house \citep{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas. With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.

To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman2013bayesian}. If these variances are similar, $\hat{R}\approx 1$, and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \citep{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that generated each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test). It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \citep{vehtari2019rank}. Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.

%$R^*$ calculation is enabled in Pints inference software \citep{Clerx2019Pints}.

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \citep{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.

@@

The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \citep{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for models with 1000 samples), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

@@

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \citep{vehtari2019rank} and two additional examples based on a multivariate normal model.

@@

First, we use the code provided in \citep{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.

@@

In GBMs, it is possible to calculate variable importance (see, for example, \citep{friedman2001greedy} and \citep{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that the variable importance measure provides information complementary to $\hat{R}$ and ESS.

@@

\subsection{Infinite variance: Cauchy example}
We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance. Like \citep{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
%
\begin{equation}
x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50,
\end{equation}
%
We call this parameterisation the ``nominal'' version of this model.

In addition, we also use Stan to sample from an ``alternative'' parameterisation of the Cauchy, based on a scale mixture of Gaussians \citep{vehtari2019rank},
%
\begin{align}
a_j \sim  \N(0,1), \qquad
b_j \sim  \text{Gamma}(0.5, 0.5), \qquad
x_j =  a_j/\sqrt{b_j}.
\end{align}
%
The distribution of the $x$ vector is the same under both parameterisations, although the thin-tailed $(a,b)$ vectors define a higher dimensional posterior that improves sampling efficiency.

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \citep{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \citep{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

@@

To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target. One measure of distributional "closeness" is the KL-divergence, which, in this case, could measure the divergence from the target to the sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated. The trouble is, for distributions like the Cauchy with fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails. This means that approximate KL-divergence is unreliable for these types of model. Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents. Specifically, we calculate the $R^2$ for the regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities. In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 dimensions.

In Fig. \ref{fig:cauchy_convergence}A, we plot this ``quantile $R^2$'' as a function of MCMC sample size for both parameterisations of the Cauchy model. This shows that after c.10,000 iterations, the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles. Next, in Fig. \ref{fig:cauchy_convergence}B and Fig. \ref{fig:cauchy_convergence}C, we plot two measures of $\hat{R}$, each calculated from splitting the 4 original chains into two equal halves. The first of these measures is the rank-normalised $\hat{R}$ \citep{vehtari2019rank}, which provides a separate measurement for each target dimension; in Fig. \ref{fig:cauchy_convergence}B, we show how the maximum of this measurement across all 50 dimensions changes with sample size. After c.550 iterations, the alternative parameterisation achieves $\hat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum $\hat{R}$ value: in both cases, these suggest convergence. The second measure is multivariate $\hat{R}$ \citep{brooks1998general}, which, like $R^*$, yields a single measurement across all dimensions; Fig. \ref{fig:cauchy_convergence}C shows how this metric changes with sample size for both Cauchy model parameterisations. After c. 1800 iterations, multivariate $\hat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\hat{R}>1.07$ indicating more samples are needed. In Fig. \ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models: these indicate that, after 25,000 iterations, $R^*\approx 1.05$ for the nominal model, and $R^*>2$ for the nominal parameterisation: both these $R^*$ values suggest lack of convergence. Finally, in Figs. \ref{fig:cauchy_convergence}E\&F, we plot the minimum across all the dimensions of tail- and bulk-ESS calculated as described in \cite{vehtari2019rank}. After c.180 iterations, the alternative parameterisation surpassed a tail-ESS of 400; after c.18,700, the nominal parameterisation did the same. Both models were quicker to pass 400 bulk-ESSs.

Comparing our ``omniscient'' measure of convergence shown in Fig. \ref{fig:cauchy_convergence}A, with the various derived measures, it is clear that all show a similar pattern: as sample size increases, the various statistics tend towards convergence. The rate at which these converge differs though, and $R^*$ (Fig. \ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to our ``omniscient'' measure.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
	\caption{\textbf{Measuring convergence with MCMC sample size for Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split $\hat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:cauchy_convergence}
\end{figure}

\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \citep{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
